Cheminformatics: The Computer Science of Chemical Discovery

### 2011-08-01 paper.tex
Added Aarons latest edits

\maketitle
\begin{abstract}
b  One of the most prominent success stories in all the sciences over the last decade has
  been the advance of bioinformatics: interdisciplinary collaboration between
 computer scientists and molecular biologists led to the sequencing of the human genome
  and many other accomplishments, since researchers connected biological
  concepts to the theory of string algorithms. However, despite this great
  success, few computer scientists are familiar with a related (and older!)
  discipline---cheminformatics, the use of computers to represent
 the structures of small molecules and analyze their properties. 
  Cheminformatics has a wide applicability,
  from the discovery of novel drug candidates to the optimization of
  certain physicochemical properties of small molecules. Until
 recently, the data and techniques of cheminformatics have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new ``miracle molecule.'' Only within the last decade---and largely because of
 chemists volunteering their time for an Open Science ``movement''---have researchers
  gained access to freely available software packages and databases of tens of
  millions of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria.
\end{abstract}

@@

Novel technologies in the life sciences churn out information at an ever
increasing rate---public data stores such as at the European Bioinformatics
Institute (EBI) contain in the order of 10 petabytes of biological information.
However, while biological information has been publicly available for decades,
the same has not been true for chemical information until very recently. It was
not until 2004 that a large public small molecule structure repository (PubChem)
became freely available. This was followed by other databases, as discussed later
in this article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field
\cite{faulon2010}.

@@

There is a considerable progress in open chemical and bioactivity databases like
ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
of chemical databases with each other remains a challenge, not only due to the
normalization problems in the chemical and bioactivity landscape, but also due
to sheer data volume. As mentioned earlier, we have to be able to
search in complex data, for example multi-labelled molecular graphs, where
graph labels can be different. Still, we need to be able to capture
molecular graphs in a risk-bounded machine encoded way.

Chemical graphs can differ based on the manner in which the structures
are drawn. Also chemical behaviors can result in multiple encodings of
``the same'' molecule and its tautomeric states, different protonation
states, or many variations of implicit chemical expert
knowledge. Checking whether two graphs represent the same chemical is
an expensive operation (the next section will discuss this in-depth);
thus, a representation of the structure that is unique and invariant
with respect to how the structure is drawn is needed for
identification. An early effort in this direction was the Simplified
Molecular Input Line Specification (SMILES), and specifically the
generation of a single canonical SMILES for a particular structure. However,
with time several
different SMILES implementations became available with no universal
standard of generating a canonical SMILES. 
%After years of research and various
%commercial products, there is today no single public standard for a
%line-notation of molecular labeled graphs, which is available to a
%broad audience (academic and
%commercial). %The very same problem does also exist for
%bioassay ontologies and the lack of options to normalize the property spaces of
%chemical compounds, e.g. biological assays or clinical data outcome
%(biobanking).

@@

Graph substructure matching is a variant of \emph{graph isomorphism detection},which is
well known to be computationally intractable [3]. To execute a graph isomorphism
search across a full chemical database of thousands or even millions of
structures is infeasible. For that reason, complex heuristics must be
used to drastically filter the candidate search space. Structure-based
\emph{fingerprints} are commonly used for this purpose. Fingerprints encode
characteristic features of a given chemical structure in a Boolean array, or
bitmap. The fingerprint is created by an algorithm which generates bitmap
patterns for each atom in the structure; then each atom and its nearest
neighbors, including the bonds between them; then each group of atoms connected
by paths of increasing lengths up to the maximum implemented for the algorithm
(commonly around eight). The bitmap patterns are added together into the final
fingerprint. If all bits in a query fingerprint are also present in the target
fingerprint of a stored database structure, only then is the structure subjected
to the computationally expensive subgraph matching algorithm. Bit operations are
very fast and independent of the number of atoms in a structure due to the fixed
length of the fingerprint.

@@

Another area where enumeration plays a role is in the \emph{generation
  of chemical reaction networks}. The problem here consists of
enumerating all the possible compounds that can be produced by
applying reaction rules to a set of initial molecules. By reversing
the reaction rules one can also find sets of starting compounds
necessary for producing a given target. This latter process is known
in chemistry as \emph{retrosynthesis}, ever since the works of
E.J. Corey, who received the Nobel Prize in Chemistry in
1990. Designing new drugs or chemicals, understanding the kinetics of
combustion and petroleum refining, studying the dynamics of metabolic
networks, applying metabolic engineering and synthetic biology to
produce heterologous compounds in microorganisms, all are applications
that require the enumeration of reaction networks. As reviewed in
Chapter 11 in \cite{faulon2010} several network enumeration techniques
have been developed. However, these techniques generally suffer from a
combinatorial explosion of intermediate compounds being produced. One
way to limit the number of compounds generated is to simulate the
dynamics of the network while it is being constructed and remove
compounds of low concentration. Following that idea, methods have been
developed based on the Gillespie Stochastic Simulation Algorithm (SSA)
to compute on-the-fly species concentrations. Chemical reaction
network enumeration and sampling is an active field of research,
particularly in the context of metabolism, either to study
biodegradation, or to propose metabolic engineering strategies to
biosynthesize compounds of commercial interest. The difficulty with
metabolic network design is that in addition to network generation
based on reactions, one also needs to verify that there are enzymatic
events possible for enabling reactions catalyses. That additional task
requires encompassing both chemical and sequence information and the
development of tools that are at the interface between
chemoinformatics and bioinformatics.

@@

First, we must consider the goal of a predictive model in a
cheminformatics setting. In some cases, pure predictive ability is
desired---such as in a virtual screening setting. For such cases, the
complexity of the modeling approach is immaterial, as it leads to
accurate and generalizable results. Here a large-scale mining capacity
is crucial. On the other hand, when working with chemists and
biologists, the interpretability of a model and risk reduction
arguments are paramount, since we have to translate better-predicted
\emph{in silico} compounds into real-world compounds. In such cases,
black box methods can be less useful, though attempts have been made
to interpret vectorial models built using neural networks, SVMs or
random forests. Related to this, there is a never-ending debate on the
choice of descriptors, \emph{e.g.}, by applying feature selection.

@@

It is worth mentioning that chemistry is traditionally a field where many people
have been used to huge bookshelves of reactions and compounds. Chemists changed
to online ELN collaboration not only to increase efficiency, but also
because it allowed collaboration with trusted colleagues. Third party paper
catalogs and academic publications proved to be
inefficient, and even more critically, they did not take compounds of trusted
colleagues into account. Overall the change of management to ELNs, similar to
``Enterprise 2.0,'' required delivering on the promise to increase
interconnectivity, serve as silo-breaker, and to improve collaboration
opportunities, and indeed it already performed these functions many years ago.

@@

\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an
industrial and academic discipline with roots dating back to the
1960s, with the creation of DENDRAL, the first expert system of any
kind.  Much of its development from the 1960s until the new millenium
was performed in secret, because of competition in the pharmaceutical
industry.  While chemists were among the first scientists to embrace
web-based tools of knowledge management, research into fundamental
algorithmic questions of structure enumeration and database management
(and a host of other topics) have suffered, because of limited
academic access to either tools or data.  As we enter the second
decade of the 21st century, however, this landscape is rapidly
shifting.  There now exist free-of-charge open-source cheminformatics
toolkits, and open databases with tens of millions of compounds.  We
believe the prospects for academic and industrial collaboration
between chemists and computer scientists are bright, and we
would like to encourage more computer scientists to participate in
cheminformatics research.  There is an admittedly steep learning
curve, as the structural issues faced by chemists are often too
detailed and subtle to admit as clean an abstraction as ``algorithms
on a string,'' the way many bioinformatics algorithms can be abstracted into a CS
framework.  Nevertheless, because of the changing landscape, there are
many research questions we are only now in position to try to answer:
for a theorist, graph-theoretic questions of three-dimensional
enumeration; for a database designer, more effective search algorithms
and ontologies; for a practical programmer, expansion of the many
open-source cheminformatics projects.  If more chemists think
algorithmically, and more computer scientists think ``chemically,''
%we may discover ``miracle drugs'' more consistently than ever before.
we are much better positioned taking not only ``simple'' chemicals
into account (only one identifier and isomer possible per molecule),
but also their complex relationships, transformations, and
combinatorial challenges. This will allow us to
normalize, search, and use all chemical information, around us and
within us.

>>>>

\maketitle
\begin{abstract}
One of the most prominent success stories in all the sciences over the last decade has
  been the advance of bioinformatics: interdisciplinary collaboration between
 computer scientists and molecular biologists led to the sequencing of the human genome
  and many other accomplishments, since researchers connected biological
  concepts to the theory of string algorithms. However, despite this great
  success, few computer scientists are familiar with a related (and older!)
  discipline---cheminformatics, the use of computers to represent
 the structures of small molecules and analyze their properties.
  Cheminformatics has a wide applicability,
  from the discovery of novel drug candidates to the optimization of
  certain physicochemical properties of small molecules. Until
 recently, the data and techniques of cheminformatics have been closely guarded
  secrets of companies whose financial success depended on being the first to produce
  the new ``miracle molecule.'' Only within the last decade---and largely because of
 chemists volunteering their time for an Open Science ``movement''---have researchers
  gained access to freely available software packages and databases of tens of
  millions of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago, but whose
  solutions are critical to research ranging from determining the behavior of small
  molecules in biological pathways, to finding a cure for malaria.
\end{abstract}

@@

Novel technologies in the life sciences churn out information at an ever
increasing rate---public data stores such as the one at the European Bioinformatics
Institute (EBI) contain on the order of 10 petabytes of biological information.
However, while biological information has been publicly available for decades,
the same has not been true for chemical information until very recently. It was
not until 2004 that a large public small molecule structure repository (PubChem)
became freely available. This was followed by other databases, as discussed later
in this article. In the same vein, open source software and the publication of
algorithms have only recently become a main focus of the field
\cite{faulon2010}.

@@

There is considerable progress in open chemical and bioactivity databases like
ChEMBL and NCBI-PubChem, which are both freely available. Still, the integration
of chemical databases with each other remains a challenge, not only due to the
normalization problems in the chemical and bioactivity landscape, but also due
to sheer data volume. As mentioned earlier, we have to be able to
search in complex data, for example multi-labelled molecular graphs, where
graph labels can be different. Still, we need to be able to capture
molecular graphs in a risk-bounded machine encoded way.

Chemical graphs can differ based on the manner in which the structures
are drawn. Also chemical behaviors can result in multiple encodings of
``the same'' molecule and its tautomeric states, different protonation
states, or many variations of implicit chemical expert
knowledge. Checking whether two graphs represent the same chemical is
an expensive operation (the next section will discuss this in-depth);
thus, a representation of the structure that is unique and invariant
with respect to how the structure is drawn is needed for
identification. An early effort in this direction was the Simplified
Molecular Input Line Specification (SMILES), and specifically the
generation of a single canonical SMILES for a particular structure. However,
over the course of time, several
different SMILES implementations became available, with no universal
standard for generating a canonical SMILES.
%After years of research and various
%commercial products, there is today no single public standard for a
%line-notation of molecular labeled graphs, which is available to a
%broad audience (academic and
%commercial). %The very same problem does also exist for
%bioassay ontologies and the lack of options to normalize the property spaces of
%chemical compounds, e.g. biological assays or clinical data outcome
%(biobanking).

@@

Graph substructure matching is a variant of \emph{graph isomorphism detection},which is
widely believed to be computationally intractable [3]. To execute a graph isomorphism
search across a full chemical database of thousands or even millions of
structures is infeasible. For that reason, complex heuristics must be
used to drastically filter the candidate search space. Structure-based
\emph{fingerprints} are commonly used for this purpose. Fingerprints encode
characteristic features of a given chemical structure in a Boolean array, or
bitmap. The fingerprint is created by an algorithm which generates bitmap
patterns for each atom in the structure; then each atom and its nearest
neighbors, including the bonds between them; then each group of atoms connected
by paths of increasing lengths up to the maximum implemented for the algorithm
(commonly around eight). The bitmap patterns are added together into the final
fingerprint. If all bits in a query fingerprint are also present in the target
fingerprint of a stored database structure, only then is the structure subjected
to the computationally expensive subgraph matching algorithm. Bit operations are
very fast and independent of the number of atoms in a structure due to the fixed
length of the fingerprint.

@@

Another area where enumeration plays a role is in the \emph{generation
  of chemical reaction networks}. The problem here consists of
enumerating all the possible compounds that can be produced by
applying reaction rules to a set of initial molecules. By reversing
the reaction rules one can also find sets of starting compounds
necessary for producing a given target. This latter process is known
in chemistry as \emph{retrosynthesis}, ever since the works of
E.J. Corey, who received the Nobel Prize in Chemistry in
1990. Designing new drugs or chemicals, understanding the kinetics of
combustion and petroleum refining, studying the dynamics of metabolic
networks, applying metabolic engineering and synthetic biology to
produce heterologous compounds in microorganisms, all are applications
that require the enumeration of reaction networks. As reviewed in
Chapter 11 in \cite{faulon2010} several network enumeration techniques
have been developed. However, these techniques generally suffer from a
combinatorial explosion of intermediate compounds being produced. One
way to limit the number of compounds generated is to simulate the
dynamics of the network while it is being constructed and remove
compounds of low concentration. Following that idea, methods have been
developed based on the Gillespie Stochastic Simulation Algorithm (SSA)
to compute on-the-fly species concentrations. Chemical reaction
network enumeration and sampling is an active field of research,
particularly in the context of metabolism, either to study
biodegradation, or to propose metabolic engineering strategies to
biosynthesize compounds of commercial interest. The difficulty with
metabolic network design is that in addition to network generation
based on reactions, one also needs to verify that there are enzymatic
events possible for enabling reactions catalyses. That additional task
requires encompassing both chemical and sequence information and the
development of tools that are at the interface between
cheminformatics and bioinformatics.

@@

First, we must consider the goal of a predictive model in a
cheminformatics setting. In some cases, pure predictive ability is
desired---such as in a virtual screening setting. For such cases, the
complexity of the modeling approach is immaterial, as it leads to
accurate and generalizable results. Here a large-scale mining capacity
is crucial. On the other hand, when working with chemists and
biologists, the interpretability of a model and risk reduction
arguments are paramount, since we have to translate better-predicted
\emph{in silico} compounds into real-world compounds. In such cases,
black box methods can be less useful, though attempts have been made
to interpret vectorial models built using neural networks, Support Vector Machines or
random forests. Related to this, there is a never-ending debate on the
choice of descriptors, \emph{e.g.}, by applying feature selection.

@@

It is worth mentioning that chemistry is traditionally a field where many people
have been used to huge bookshelves of reactions and compounds. Chemists changed
to online ELN collaboration not only to increase efficiency, but also
because it allowed collaboration with trusted colleagues. Third party paper
catalogs and academic publications proved to be
inefficient, and even more critically, they did not take compounds of trusted
colleagues into account. Overall the change of management to ELNs, similar to
``Enterprise 2.0,'' required delivering on the promise to increase
interconnectivity, and to improve collaboration
opportunities, and indeed it already performed these functions many years ago.

@@

\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an
industrial and academic discipline with roots dating back to the
1960s, with the creation of DENDRAL, the first expert system of any
kind.  Much of the development of cheminformatics from the 1960s until the new millenium
was performed in secret, because of competition in the pharmaceutical
industry.  While chemists were among the first scientists to embrace
web-based tools of knowledge management, research into fundamental
algorithmic questions of structure enumeration and database management
(and a host of other topics) have suffered, because of limited
academic access to either tools or data.  As we enter the second
decade of the 21st century, however, this landscape is rapidly
shifting.  There now exist free-of-charge open-source cheminformatics
toolkits, and open databases with tens of millions of compounds.  We
believe the prospects for academic and industrial collaboration
between chemists and computer scientists are bright, and we
would like to encourage more computer scientists to participate in
cheminformatics research.  There is an admittedly steep learning
curve, as the structural issues faced by chemists are often too
detailed and subtle to admit as clean an abstraction as ``algorithms
on a string,'' the way many bioinformatics algorithms can be abstracted into a CS
framework.  Nevertheless, because of the changing landscape, there are
many research questions we are only now in position to try to answer:
for a theorist, graph-theoretic questions of three-dimensional
enumeration; for a database designer, more effective search algorithms
and ontologies; for a practical programmer, expansion of the many
open-source cheminformatics projects.  If more chemists think
algorithmically, and more computer scientists think ``chemically,''
%we may discover ``miracle drugs'' more consistently than ever before.
we are much better positioned taking not only ``simple'' chemicals
into account (only one identifier and isomer possible per molecule),
but also their complex relationships, transformations, and
combinatorial challenges. This will allow us to
normalize, search, and use all chemical information, around us and
within us.
